---
title: "Introduction"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(readxl)
knitr::opts_chunk$set(echo = FALSE)
```


## first steps with R

### Exercise 1

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise 2

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

### Exercise 3

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

### Quiz

Some questions to verify that you know the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which packages contain functions for installing other R packages?",
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("remote", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("ggplot2", correct = TRUE)
  )
)
```


## explore the data

### read data from XLSX files

Read a file located at `inst/extdata/data_C202.xlsx`. You need the `readxl` package; here it is already pre-loaded.
```{r read-xlsx, exercise=TRUE}

```
