---
title: "Introduction"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(readxl)
knitr::opts_chunk$set(echo = FALSE)
```

This tutorial is one of the four created in May 2024 by [Giovanni Bonafè](https://www.linkedin.com/in/giovannibonafe/) for the __6th Workshop on Water Resources in Developing Countries__ at ICTP in Trieste, Italy. Source code is available [here](https://github.com/jobonaf/tutorialhydrology).

I suggest you do the tutorials in this order:

1. [introduction](https://jobonaf.shinyapps.io/tutorialhydrology-intro/) to R and to the hydrological data 
2. [plot](https://jobonaf.shinyapps.io/tutorialhydrology-plot/) the hydrological data 
3. [fit](https://jobonaf.shinyapps.io/tutorialhydrology-fit/) the rating curve 
4. [estimate](https://jobonaf.shinyapps.io/tutorialhydrology-estimate/) the discharge using the fitted rating curve 


## first steps with R

### Exercise 1

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise 2

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}

add <- function() {
  
}

```

### Exercise 3

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}

mtcars

```

```{r print-limit-hint}
head(mtcars, n=5)
```

### Quiz

```{r quiz-1}
quiz(
  question("Which packages contain functions for installing other R packages?",
           answer("tools"),
           answer("utils", correct = TRUE),
           answer("remote", correct = TRUE),
           answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
           answer("lattice", correct = TRUE),
           answer("tools"),
           answer("stats"),
           answer("ggplot2", correct = TRUE)
  ),
  question("Which of the following assign value 5 to object A?",
           answer("A = 2 + 3", correct = TRUE),
           answer("2 + 3 = A"),
           answer("2 + 3 -> A", correct = TRUE),
           answer("A <- 2 + 3", correct = TRUE)
  ), caption=""
)
```


## arrange the data

### read data from XLSX files

Try reading the second sheet of a file located at `data/data_C202.xlsx`, skipping the first line. You need the `readxl` package. 

```{r read-xlsx, exercise=TRUE}

library(readxl)
dat <- read_xlsx("data/data_C202.xlsx", sheet=1, skip=0)
dat

```

### explore the dataset

If you need a summary of the dataset, functions `dim`, `str` and `summary` are good choices. Here you can try them. Consider that metadata and data are pre-loaded in objects `anag` and `dat`.

```{r prepare-data, include=FALSE}

anag <- read_xlsx("data/data_C202.xlsx", sheet=1, skip=1)
dat <- read_xlsx("data/data_C202.xlsx", sheet=2, skip=1)

```

```{r data-summary, exercise=TRUE, exercise.setup = "prepare-data"}

str(dat)

```

Now you should be able to answer to a couple of questions

```{r quiz-2}
quiz(
  question("What is the maximum stage detected by the sensor (cm)?",
    answer(28),
    answer(102, correct = TRUE),
    answer(535)
  ),
  question("What is the average temperature of the water (°C)?",
    answer(10.76, correct = TRUE),
    answer(10),
    answer(14)
  ),
  question("In which municipality is located the station?",
    answer("Tagliamento"),
    answer("Cleulis"),
    answer("But"),
    answer("Paluzza", correct = TRUE)
  ), caption=""
)
```

