---
title: "Plot the data"
author: G.Bonafè
date: 2024-05-23
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
library(dplyr)
library(lubridate)
library(tidyr)
load("data/obs_discharge.rda")
knitr::opts_chunk$set(echo = FALSE)
```

This tutorial is one of the four created in May 2024 for the __6th Workshop on Water Resources in Developing Countries__ at ICTP in Trieste, Italy. Source code is available [here](https://github.com/jobonaf/tutorialhydrology).

I suggest you do the tutorials in this order:

1. [introduction](https://jobonaf.shinyapps.io/tutorialhydrology-intro/) to R and to the hydrological data 
2. [plot](https://jobonaf.shinyapps.io/tutorialhydrology-plot/) the hydrological data 
3. [fit](https://jobonaf.shinyapps.io/tutorialhydrology-fit/) the rating curve 
4. [estimate](https://jobonaf.shinyapps.io/tutorialhydrology-estimate/) the discharge using the fitted rating curve 


## Explore a larger dataset

Now we have a larger dataset, named `obs_discharge`. It has been built by merging and rearranging data from 19 stations, in the same format as the XLSX file we have already seen in the first tutorial [Introduction](https://jobonaf.shinyapps.io/tutorialhydrology-intro/).
Let's explore the dataset with functions `colnames`, `dim`, `str` and `summary`.

```{r explore-data, exercise=TRUE}

colnames(obs_discharge)

```

With package `dplyr` (here pre-loaded) you can `filter`, `group_by` and `summarize` the data. These and other functions can be concatenated with the pipe operator `%>%`.

```{r filter-data, exercise=TRUE}

obs_discharge %>%
  filter(Variable=="Q") %>%
  group_by(River, StationCode) %>%
  summarize(v=max(Value))

```

Try to answer to these questions. Consider adding `%>% arrange(desc(v))` in order to sort the final table by column `v`. 

```{r quiz-1}
quiz(
  question("Which is the station where the maximum discharge has been measured?",
           answer("Fiume Isonzo"),
           answer("N021", correct = TRUE),
           answer("NO21")
  ),
  question("Which is the station where the maximum stage has been measured by the sensor?",
           answer("Fiume Meduna"),
           answer("N021"),
           answer("A323", correct = TRUE)
  )
  , caption=""
)
```

## Time series plots

Here we plot the time series of the measured data. First we choose the discharge `Q` of the station `N021`.

```{r timeseries-1, exercise=TRUE}

dat <- obs_discharge %>% filter(Variable=="Q", StationCode=="N021")
ggplot(data = dat, mapping = aes(x=Date, y=Value)) +
  geom_point()

```

Now try with other variables and other stations. You can change or add other geometries (`geom_line`, `geom_col`).
You can also plot two or more variables (or stations) and use the `color` to distinguish them from each other.

```{r timeseries-2, exercise=TRUE}

dat <- obs_discharge %>% filter(Variable %in% c("H","Q"), StationCode=="N021")
ggplot(dat, aes(x=Date, y=Value, color=Variable)) +
  geom_point()

```
Another way to distinguish the variables or the stations are the `facet_*` functions. Here you see an example. 

```{r timeseries-3, exercise=TRUE}

dat <- obs_discharge %>% filter(StationCode=="N021")
ggplot(data = dat, mapping = aes(x=Date, y=Value)) +
  facet_wrap("Variable")+
  geom_point()

```

```{r quiz-2}
quiz(
  question("How can you rescale the y axis of each panel according to the values of each variable, keeping the x scale the same for each panel?",
           answer("scale='free'"),
           answer("scale='free_y'", correct = TRUE),
           answer("free_y=TRUE")
  )
  , caption=""
)
```

## Box and whiskers plots

Box and whiskers plots are used to show distributions of numeric data values, especially when you want to compare them between multiple groups. 

```{r boxplot-1, exercise=TRUE}

dat <- obs_discharge %>% filter(Variable=="Q")
ggplot(dat, aes(y=StationCode, x=Value)) +
  geom_boxplot()

```

With function `mutate` you can add a new column to your dataset. With package `lubridate` (here pre-loaded) you have function `month`, so you can add the month to the dataset. Now plot the new dataset, using the month as `group`, so that you represent the annual course of the observed discharge.

```{r boxplot-2, exercise=TRUE, exercise.eval=TRUE}

dat <- obs_discharge %>% 
  filter(Variable=="Q", StationCode=="N021") %>%
  mutate(Month=month(Date))
dat

```

```{r boxplot-2-hint}

# add the following lines:
ggplot(dat, aes(x=Month, y=Value, group=Month)) +
  geom_boxplot() +
  scale_x_continuous(breaks=1:12, label=month.abb)

```

### Exercise

Compare the annual course of different stations and/or of different variables. Give your own interpretation to the results.

## Reshape the data

In our dataset each row contains the value of a different variable ("long" or ["narrow" format](https://en.wikipedia.org/wiki/Wide_and_narrow_data#Narrow)).

: Example of "long" table

|Date | Station | Variable | Value|
|:--|:--|:--|:--|
|12/3 | A1      | Q        | 23.1|
|12/3 | A1      | H        | 1.7|
|12/3 | A1      | T        | 8.1|

You may want to compare two variables with a scatter plot, so you need to reshape the dataset in order to get one distinct column for each variable (["wide" format](https://en.wikipedia.org/wiki/Wide_and_narrow_data#Wide)). 

: Example of "wide" table

|Date | Station | Q    | H    | T    |
|:--|:--|:--|:--|:--|
|12/3 | A1      | 23.1 | 1.7  | 8.1  |

To achieve this goal, function `pivot_wider` of package `tidyr` is right for you.


```{r reshape, exercise=TRUE, exercise.eval=TRUE}

obs_discharge %>% 
  pivot_wider(id_cols = c(TimeStart, StationCode), 
              names_from = Variable, 
              values_from = Value)

```

## Scatter plot

Now, after reshaping the data, you have a `data.frame` with one distinct column for each measured variable. On a single row you have simultaneous measures. So you should be able to build a scatter plot with discharge `Q` vs stage `H`. In the Cartesian plane a point will represents a pair _(H,Q)_.

```{r scatter, exercise=TRUE}

obs_discharge %>% 
  pivot_wider(id_cols = c(TimeStart, StationCode), 
              names_from = Variable, 
              values_from = Value) -> dat

```

```{r scatter-hint}

obs_discharge %>% 
  filter(Variable %in% c("Q","H"), StationCode %in% c("N021","A323"))%>% 
  pivot_wider(id_cols = c(TimeStart, StationCode), 
              names_from = Variable, 
              values_from = Value) -> dat
ggplot(dat, aes(x=H, y=Q, col=StationCode)) +
  geom_point()

```

