---
title: "Fit a rating curve"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(bdrc)
load("data/obs_discharge.rda")
load("data/obs_n105.rda")
knitr::opts_chunk$set(echo = FALSE)
```

This tutorial is one of the four created in May 2024 by [Giovanni Bonafè](https://www.linkedin.com/in/giovannibonafe/) for the __6th Workshop on Water Resources in Developing Countries__ at ICTP in Trieste, Italy. Source code is available [here](https://github.com/jobonaf/tutorialhydrology).

I suggest you do the tutorials in this order:

1. [introduction](https://jobonaf.shinyapps.io/tutorialhydrology-intro/) to R and to the hydrological data 
2. [plot](https://jobonaf.shinyapps.io/tutorialhydrology-plot/) the hydrological data 
3. [fit](https://jobonaf.shinyapps.io/tutorialhydrology-fit/) the rating curve 
4. [estimate](https://jobonaf.shinyapps.io/tutorialhydrology-estimate/) the discharge using the fitted rating curve 

## fit a curve on a scatter plot

We have already seen how to prepare the data so that it is suitable for a scatter plot. In this example we see three stations. We distinguish them using different colors. Use `geom_smooth` to add a curve fitting the data.

```{r scatter, exercise=TRUE, exercise.eval=TRUE}

obs_discharge %>% 
  filter(Variable %in% c("Q","H"), StationCode %in% c("N021","A323","C503"))%>% 
  pivot_wider(id_cols = c(TimeStart, StationCode), 
              names_from = Variable, 
              values_from = Value) -> dat
ggplot(dat, aes(x=H, y=Q, col=StationCode)) +
  geom_point()

```

```{r scatter-hint}
obs_discharge %>% 
  filter(Variable %in% c("Q","H"), StationCode %in% c("N021","A323","C503"))%>% 
  pivot_wider(id_cols = c(TimeStart, StationCode), 
              names_from = Variable, 
              values_from = Value) -> dat
ggplot(dat, aes(x=H, y=Q, col=StationCode)) +
  geom_point()+
  geom_smooth(method="gam")
```

### exercise 1

Test different `method`s for `geom_smooth`. Which ones seem the most suitable to you?

### exercise 2

Using `facet_wrap('StationCode')` instead of the color to distinguish the stations, you can plot them all together. Given the variability of H and Q between the various stations, make an appropriate choice for the `scale` parameter. Which stations would you __exclude__ from a rating curve fitting procedure based on this data? why?

Alternatively, which pre-processing of the data would you suggest?

## fit a rating curve

We will use R package `bdrc` to fit a rating curve with a bayesian approach. That's quite expensive in terms of computing resources, so if you do this tutorial interactively on shinyapps.io, please focus only on station _N105_, so as not to exceed the limit of available computing resources. The `data.frame` `obs_n105` is ready for this purpose.


```{r fit-1}

fitted <- gplm(Q~H, data=obs_n105) 

```
